---
title: "PSTAT126-FINAL"
output:
  pdf_document: null
  latex_engine: xelatex
  html_document:
    df_print: paged
---

``` {r, echo = F, message = F}
library(skimr)
library(dplyr)
library(ggplot2)
```
## PART - 1 Data Description and Descriptive Statistics

### Description of Variables and Observational Unit 

The **Color** variable is a color grading scale that ranges from D to H (in the data set) and rates the colorless or
near colorless of a white diamond. Diamonds in the ranges of D to F are the best that contain no other color and are
of the highest quality. Diamonds in the ranges of G to J are still high quality but contain slight traces of color.

The **Clarity** variable rates the presence of internal or external blemishes, known as inclusions within the
diamond. In the data set the ratings go from VVS2, VS1, VS2, SI1, SI2. VVS2 means that inclusions are very difficult
to see under 10x magnification. VS means that inclusions are difficult to see under 10x magnification. VS2 means
that inclusions are somewhat easier to see under 10x magnification, but still minor. SI1 means that inclusions are
noticeable under 10x magnification. SI2 means that inclusions are more easily noticeable under 10x magnification.

The **Price** variable determines the monetary value of a diamond. There are a lot of factors that go into
determining the value of a diamond like cut, clarity, color, shape, and many more. 

The **Table** of a diamond refers to the large flat face when looking down at the top of the diamond. In the data
set the variable is a percentage of how big the table is compared to the total diameter of the diamond.

The **Depth** of a diamond is the distance from the large flat surface of the top of the diamond, known as the
table, to the bottom of the diamond, known as the culet. In the data set the variable is a percentage of how deep a
diamond in relation to its width. 

The **Observational Unit** of the data set is a single diamond out of the total 54,000 diamonds that make up the data
set. Each contain the variables stated before as well as 6 other variables that are outside the scope of this
project.

### Summary of The Data

``` {r Part 1.0, echo = T, results = T}
dataset <- read.csv("/users/robeh/desktop/DiamondsPrices2022.csv")
set.seed(1)
diamonds_subset <- sample_n(dataset, 500)

table(diamonds_subset$color)
prop.table(table(diamonds_subset$color))
table(diamonds_subset$clarity)
prop.table(table(diamonds_subset$clarity))

summary(diamonds_subset$price)
summary(diamonds_subset$table)
summary(diamonds_subset$depth)
```

The summary statistics you see come form a subset of the Diamonds data set, which we created by taking at random 500
observations to create this subset, we can see the summaries of the 5 variables that are within the scope of the
experiment being conducted.

Looking at the variable **Color** we see that it's divided into 7 categories that represent the diamond's rating for
color. The rating **G** has the most with 109 diamonds with that rating whereas the rating **I** has the least amount
of diamonds at just 35. Looking at it from a percentage point of view, since there are 7 categories and if they were evenly distributed among the categories, each would have about 14.3%, which most of the categories are close to.

On to the second categorical variable, **Clarity** the data shows out of the 8 categories the 2 with the most
observations were **VS2** and **SI1** which respectively had 121 and 120 of the observations. The category with least 
observations by far was **I1** which had a total of 6 observations. If the observations were perfectly distributed
among the 8 categories they would each have 12.5%, however in this summary we see that many are far off from that 
percentage and vary a lot.

Moving onto **Price**, the mean price of a diamond was \$4,249 with the minimum price and maximum price being \$362 and 
\$18,757 respectively. When it comes to the quartiles 25% of the diamonds were under \$917, half of the diamonds were
under \$2,657 and 75% of the diamonds were under \$5374. Looking at these results we see how the mean and median are
around \$2,400 off from each other, meaning that there were a few numbers that were very high skewing the average. This
is shown even more when you look at the 75% quartile and there is over a \$13,000 difference between them. 

Looking at the **Table** variable which are percentages in the data set, it shows that the mean is 57.37% and the 
minimum and maximum of the table of the diamond is 52% and 66%. With the median being very close to the average sitting
at exactly 57% we know that there aren't to many far off table observations. However, looking at the difference of the 
maximum percent and the 75% quartile there is 7% difference which is a lot, considering that the total dispersion
from the minimum and maximum is 14% which makes the 7% difference half of the whole dispersion. 

Lastly, the **Depth** which are also percentages in the data set. The summary shows that the mean is 61.87% and the 
minimum and maximum percentages are 56.9% and 70.2% respectively. When it comes to the quartiles, 25% of the depths are
under 56%, half of the depths are under 61.9%, and 75% of the depths are under 62.6%. Something interesting to note is that similarly to the **Table** variable, the difference between the max value and 75% quartile value was 7.6% which is
actually more than half of the difference between the minimum and maximum value which was 13.3%. 

``` {r Part 1.1, echo = F, results = T}
table(diamonds_subset$color, diamonds_subset$clarity)

ggplot(diamonds_subset, aes(x = table, y = price)) + geom_point() + labs(title = "Comparing Price($) and Table(%)", x = "Table(%)", y = "Price($)") 

ggplot(diamonds_subset, aes(x = depth, y = price)) + geom_point() + labs(title = "Comparing Price($) and Depth(%)", x = "Depth(%)", y = "Price($)") 
```

``` {r Part 1.2, echo = F, results = F}
set.seed(2)
diamonds_subset_2 <- sample_n(dataset, 500) 

table(diamonds_subset_2$color)
table(diamonds_subset_2$clarity)
summary(diamonds_subset_2$price)
summary(diamonds_subset_2$table)
summary(diamonds_subset_2$depth)
```

``` {r Part 1.3, echo = F, results = F}
set.seed(3)
diamonds_subset_3 <- sample_n(dataset, 500)

table(diamonds_subset_3$color)
table(diamonds_subset_3$clarity)
summary(diamonds_subset_3$price)
summary(diamonds_subset_3$table)
summary(diamonds_subset_3$depth)
```




