---
title: "PSTAT126-Final-V2"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2024-06-07"
---

## PART - 1 Data Description and Descriptive Statistics

```{r Part 1.0, echo = T, results = T}
Diamond_dataset <- read.csv("/users/robeh/desktop/DiamondsPrices2022.csv")
attach(Diamond_dataset)
head(Diamond_dataset)

library(dplyr)
set.seed(1)
diamonds_sample <- sample_n(Diamond_dataset, 500)

library(skimr)
# Have to use this instead of skim, because I get error trying to load the histogram 
# due to a latex error
skim_without_charts(diamonds_sample)
```

### Description of Variables and Observational Unit 

Main Variables Being Testing:

The **Color** variable is a color grading scale that ranges from D to H (in the data set) and rates the colorless or
near colorless of a white diamond. Diamonds in the ranges of D to F are the best that contain no other color and are
of the highest quality. Diamonds in the ranges of G to J are still high quality but contain slight traces of color.

The **Clarity** variable rates the presence of internal or external blemishes, known as inclusions within the
diamond. In the data set the ratings go from VVS2, VS1, VS2, SI1, SI2. VVS2 means that inclusions are very difficult
to see under 10x magnification. VS means that inclusions are difficult to see under 10x magnification. VS2 means
that inclusions are somewhat easier to see under 10x magnification, but still minor. SI1 means that inclusions are
noticeable under 10x magnification. SI2 means that inclusions are more easily noticeable under 10x magnification.

The **Price** variable determines the monetary value of a diamond. There are a lot of factors that go into
determining the value of a diamond like cut, clarity, color, shape, and many more. 

The **Table** of a diamond refers to the large flat face when looking down at the top of the diamond. In the data
set the variable is a percentage of how big the table is compared to the total diameter of the diamond.

The **Depth** of a diamond is the distance from the large flat surface of the top of the diamond, known as the
table, to the bottom of the diamond, known as the culet. In the data set the variable is a percentage of how deep a
diamond in relation to its width. 

The **Observational Unit** of the data set is a single diamond out of the total 54,000 diamonds that make up the data
set. Each contain the variables stated before as well as 6 other variables that are outside the scope of this
project.

Other Variables in The Data Set:

The **Carat** refers to the weight unit that measures gemstones, in this case it is measuring diamonds. One carat is 
equal to 200 milligrams.

The **Cut** variable is a categorical variable is a rating that is an important piece of the puzzle that determines how
good the diamond is overall. It goes into how the diamond looks visually as well as how well the diamond is crafted.

The **X**, **Y**, and **Z** variable just refers to the dimensions of the diamond. The **x** is the length in mm, the 
**y** is the width in mm, and the **z** is the depth in mm. 

```{r Part 1.2, echo = T, results = T}
fit0 <- lm(price ~ table + clarity, data = diamonds_sample)
summary(fit0)
```

From the results we can see the the variables clarity and table don't have a significant effect on the price variable.
When looking at the R-Squared value, we see that they only account for 6.14% of the variance in price. 

```{r Part 1.3, echo = T, results = T}
table(diamonds_sample$color)

diamonds_sample$D = ifelse(diamonds_sample$color == "D", 1, 0)
diamonds_sample$E = ifelse(diamonds_sample$color == "E", 1, 0)
diamonds_sample$F = ifelse(diamonds_sample$color == "F", 1, 0)
diamonds_sample$G = ifelse(diamonds_sample$color == "G", 1, 0)
diamonds_sample$H = ifelse(diamonds_sample$color == "H", 1, 0)
diamonds_sample$I = ifelse(diamonds_sample$color == "I", 1, 0)
diamonds_sample$J = ifelse(diamonds_sample$color == "J", 1, 0)
attach(diamonds_sample)

fit2 <- lm(price ~ depth + E + F + G + H + I + J, data = diamonds_sample)
summary(fit2)
```

Now looking at depth and color (using dummy variables) we can see how alike the last test they have a very small 
significance on price. Back to the R-Squared value it shows that they only account for 7.47% of the variance in 
the price variable.

```{r Part1.4, echo = T, results = T}
fit3 <- lm(price ~ color + clarity + depth + table, data = diamonds_sample)
summary(fit3)
```

Constructing a model with all the variables that I randomly chose, it is apparent that the ones I chose don't have that
big of an impact when it comes to the effect of the price. In the last summary, it does show a lot of significance stars
but in the end the R-Squared value only comes out to 0.1434, something I'll have to work on from here on out on this 
project. 

### Comments on Part 1

I was surprised that the variables I choose to test showed that they didn't have that big an effect on the overall
price of the diamond. The variables I chose had a very low R-Squared value which I wasn't really expected. However, 
after looking at the data again I realized that the percentages of the depth and table don't can't really quantify the 
diamond. They give a percentage of the proportion, but diamond that are way different in size could still have the same
depth or table, so it makes sense that these values don't really have a big affect on the price. 

## PART - 2 Simple Linear Regression

```{r Part 2, echo = T, results = T}
SLR <- lm(price ~ table, data = diamonds_sample)
par(mfrow=c(2,2))
plot(SLR)
summary(SLR)

SLR1 <- lm(log(price) ~ table, data = diamonds_sample)
par(mfrow=c(2,2))
plot(SLR1)
summary(SLR1)

SLR2 <- lm(price ~ log(table), data = diamonds_sample)
par(mfrow=c(2,2))
plot(SLR2)
summary(SLR2)
```

```{r Part 3, echo = T, results = T}
anova(fit0)
```
